Quantifying the importance of a tree prior in Bayesian phylogenetics

### 2019-04-09 article.tex

\usepackage{listings}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{verbatim}
\usepackage{pgf}
\usepackage{bm}
\usepackage{multirow}
\usepackage{amsfonts}
\usepackage{array}
\usepackage{array}
\usepackage{booktabs}
\newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\usepackage{longtable}
\usepackage{rotating}
\usepackage{mathtools}

@@

  \textbf{1. }
    BEAST2 is a popular Bayesian phylogenetics software tool,
    that takes an alignment of characters (usually nucleotides) 
    and an inference model to create a
    posterior of jointly-estimated phylogenies and model parameters. 
    An important ingredient of BEAST2 is the species tree prior, 
    due to which the tool can be used for inference of 
    macroevolutionary processes underlying the species tree. 
    However, only models that allow for fast computation of 
    the prior probability are possible in practice. 
    This begs the question of how accurate the tree estimation is 
    when the real macroevolutionary processes are different 
    from those assumed in BEAST2. \\
  \textbf{2. }
    Here we present \verb;pirouette;, 
    a free, libre and open-source R package that assesses 
    the inference error BEAST2 makes based on a 
    phylogeny from any macroevolutionary diversification model. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples. \\
\end{abstract}

@@

BEAST [\cite{drummond2007beast}] is a Bayesian phylogenetics tool, 
that needs both data and priors to infer a posterior.
Of these priors, the tree prior describes how diversification occurs.
BEAST has built-in tree priors such as the Yule [\cite{yule}] and 
(constant-rate) birth-death [\cite{nee1994reconstructed}] model.
These simple tree priors are most commonly used, as these 
have sufficient biological complexity, while being computationally fast.
BEAST's successor, BEAST2 [\cite{bouckaert2014beast}],
has a package manager, that allows third-party users to extend existing 
functionality.
By writing a BEAST2 package, novel diversification models can be added.
The key ingredient of such a package would be the likelihood equation
of a phylogeny for the novel diversification model.
There have been plenty of these models (and their associated probability 
algorithms) 
developed, among others, those in which diversification is 
time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
or diversity-dependent [\cite{etienne2011diversity}],
or that can shift [\cite{etienne2012conceptual}, 
\cite{rabosky2014automatic}, \cite{alfaro2009nine}, \cite{laudanno2018sls}].
Other birth-death models treat speciation as a process that takes 
time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}], 
or that happens in bursts of simultaneous 
events [\cite{laudanno2018mbd}], or that
depends on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
or even allows concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].
Only few of these diversification models are available as a BEAST2 package.

@@

@@

Now we can formulate the first example research question that \verb;pirouette; 
can answer:

@@

In this example we use a standard generative tree prior $\mathit{P_{0}}$, 
namely the Yule (pure-birth) tree prior. 
We choose to use a small tree with six taxa to keep
the calculations short and the figure more readable.
We pick a crown age of ten time units. This value is 
completely arbitrary, but does tie in with the mutation rate 
used in simulating an alignment in the next step.

@@

\begin{lstlisting}[
  language=R, floatplacement=ht, frame=single,
  label = {lst:create_error_measure_params},
  caption = Calling \texttt{create\_error\_measure\_params}
]
error_measure_params <- create_error_measure_params(
  burn_in_fraction = 0.1, 
  error_function = get_nltt_error_function()
)
\end{lstlisting}

@@

\begin{lstlisting}[
  language=R, floatplacement=ht, frame=single,
  label = {lst:create_pir_params},
  caption = Calling \texttt{create\_pir\_params}
]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  error_measure_params = error_measure_params,
  twinning_params = NA
)
\end{lstlisting}

@@

\begin{lstlisting}[
  language=R, floatplacement=ht, frame=single,
  label = {lst:pir_run},
  caption = Calling \texttt{pir\_run}
]
errors <- pir_run(
  phylogeny = phylogeny,
  pir_params = pir_params
)
\end{lstlisting}

@@

\begin{lstlisting}[
  language=R, floatplacement=ht, frame=single,
  label = {lst:pir_plot},
  caption = Calling \texttt{pir\_plot}
]
pir_plot(errors)
\end{lstlisting}

@@

The error distribution obtained is shown as a violin plot
in figure~\ref{fig:example_1}
\richel{
  I suggest to use a histogram (or density plot) instead:
  I like how the error distribution in figure 1 looks.
}
.

@@

By doing so, this error distribution will serve as a control,
as it can be seen as the baseline error caused by stochasticity.
If the two error distributions obtained from $\mathit{P_{0}}$ and $\mathit{P}$ 
will appear similar, the error induced by the choice of a non-standard 
speciation process will be of little influence.
Given the necessity of such comparison,
we made it an inherent part of \verb;pirouette;. 
We call it 'twinning' and we use it in ERQ 3.

@@

@@

If the user has sufficient reasons to believe that one standard tree prior 
could be more suitable than others for the inference, they can set it as part 
of the inference model and to adopt site and clock models used to simulate the 
alignment. This option can be specified selecting "model\_type" to be 
"generative" as part of the inference conditions (as seen in ERQ 1, 
table~\ref{tab:RQ1}).

We show the procedure using \verb;pirouette; to answer the second research 
question:

@@

Here we start with a tree generated from an unknown 
diversification model, that has six taxa and a crown age of ten:

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_2/errors.png}
  \caption{
    Example 2: the error BEAST2 makes from a phylogeny when
    picking the best inference model.
  }
  \label{fig:example_2}
\end{figure}

@@

This time, however, we are interested in measuring both the inference error
and a baseline error. We do so by setting the twinning parameters $\mathit{T}$ to create a twin tree. A twin tree
has the same topology as the given tree, yet its branching times are simulated 
according to the standard tree prior used by BEAST2 to produce the posterior 
for the main pipeline. 
Since there is coherence between the inference model and the generative model, 
we expect BEAST2 to produce a posterior of phylogenies that are more similar to the 
original one, respect to the outcome of the main pipeline. 
Due to this, the twinning process is used to provide a control for the main 
experiment.

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_3/errors.png}
  \caption{
    Example 3: the error made by BEAST2 from a phylogeny, 
    when hand-picking a generative tree prior versus the control.
    Here, the 'twin' column shows the baseline inference error,
    as measured on the twin tree
  }
  \label{fig:example_3}
\end{figure}

@@

To answer ERQ 2 we used an evidence-based model selection procedure to obtain 
an 
error distribution.
However, it is unclear to what extent this error is due to the mismatch between 
the tree priors (from generative and inference models) or due to stochasticity. 
As mentioned in ERQ 1, \verb;pirouette; provides a tool to evaluate this, using 
the twinning process. We now apply it to the same pipeline followed in ERQ 2.

@@

Here, as in examples 2 and 3, we start with the phylogeny 
generated by an unknown speciation model (listing~\ref{lst:unknown_phylogeny}). 
The first step in \verb;pirouette; is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment}.

@@

The output is shown in figure~\ref{fig:example_4}. The alignment
generated for the true tree had most evidence for the JC site model,
strict clock model and Yule tree prior, 
with a weight of 0.981 (as 
shown in table \ref{tab:evidences_example_4}). 
The alignment
generated for the twin tree had most evidence for the JC site model,
RLN clock model and CBS tree prior,
with a weight of 0.707 (as 
shown in table \ref{tab:evidences_example_4}).

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_4/errors.png}
  \caption{
    Example 4: the error made by BEAST2 from a phylogeny, when the generative 
    model is known, compared to candidates with similar inference models, in 
    relationship to the background error.
  }
  \label{fig:example_4}
\end{figure}

@@

In ERQs 1 and 2 we used \verb;pirouette; to  
run the analysis on the main pipeline, either using a 'generative' or a 
'best\_candidate' inference model, resulting in one error distribution,
that does not mean much on its own.
In ERQs 3 and 4 we measured the baseline error caused by stochasticity,
by running a parallel twin pipeline. This yielded two error distributions:
the baseline error and the complete inference error.

@@

In this example, we use a tree generated by the Yule process.
Instead of using the common Yule tree as in example 1, we will use
a Yule tree of low (actually: zero) likelihood: 
the one as used in ERQs 2-4, by using the code shown 
in listing~\ref{lst:unknown_phylogeny}.
The first step in \verb;pirouette; is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment}.

@@

In the second step we state our experiments. 
In this example, we must both specify the generative and candidate experiments. 
Here we specify the generative model:

@@

We choose again to use the default error measurement parameters
as created in listing~\ref{lst:create_error_measure_params}
and combine this with the same twinning parameters 
as used in listing~\ref{lst:create_twinning_params}:

@@

The output is shown in figure~\ref{fig:example_5}.

The alignment
generated for the true tree had most evidence for the JC site model,
strict clock model and Yule tree prior, 
with a weight of 0.981 \richel{again} (as 
shown in table \ref{tab:evidences_example_5}). 
The alignment
generated for the twin tree had most evidence for the JC site model,
RLN clock model and CBS tree prior,
with a weight of 0.707 \richel{again} (as 
shown in table \ref{tab:evidences_example_5_twin}).

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_5/errors.png}
  \caption{
    Example 5: the error made by BEAST2 from a phylogeny, 
    when the generative model is known, 
    compared to candidates with similar inference models, 
    in relationship to the background error.
  }
  \label{fig:example_5}
\end{figure}

>>>>

\usepackage{listings}
\usepackage{hyperref}
\usepackage{todonotes}
\usepackage{verbatim}
\usepackage{pgf}
\usepackage{bm}
\usepackage{multirow}
\usepackage{amsfonts}
\usepackage{array}
\usepackage{array}
\usepackage{booktabs}
\newcolumntype{C}[1]{>{\centering\arraybackslash}p{#1}}
\newcolumntype{L}[1]{>{\raggedright\arraybackslash}p{#1}}
\usepackage{longtable}
\usepackage{rotating}
\usepackage{mathtools}
% \usepackage{float}

@@

  \textbf{1. }
    BEAST2 is a popular Bayesian phylogenetics software tool,
    that takes an alignment of characters (usually nucleotides) 
    and an inference model to create a
    posterior of jointly-estimated phylogenies and model parameters. 
    An important ingredient of BEAST2 is the species tree prior, 
    due to which the tool can be used for the inference of 
    macroevolutionary processes underlying the species tree. 
    However, only models that allow for fast computation of 
    the prior probability are possible in practice. 
    This begs the question of how accurate the tree estimation is 
    when the real macroevolutionary processes are different 
    from those assumed in BEAST2. \\
  \textbf{2. }
    Here we present \verb;pirouette;, 
    a free, libre and open-source R package that assesses 
    the inference error BEAST2 makes based on a 
    phylogeny from any macroevolutionary diversification model. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    Last, we discuss the results obtained by the examples. \\
\end{abstract}

@@

BEAST [\cite{drummond2007beast}] is a Bayesian phylogenetics tool, 
that needs both data and priors to infer a posterior.
Of these priors, the tree prior describes how diversification occurs.
BEAST has built-in tree priors such as the Yule [\cite{yule}] and 
(constant-rate) birth-death [\cite{nee1994reconstructed}] model.
These simple tree priors are most commonly used, as these 
have sufficient biological complexity, while being computationally fast.
BEAST's successor, BEAST2 [\cite{bouckaert2014beast}],
has a package manager, that allows third-party users to extend existing 
functionalities.
By writing a BEAST2 package, novel diversification models can be added.
The key ingredient of such a package would be the likelihood equation
of a phylogeny for the novel diversification model.

\iffalse
There have been plenty of these models (and their associated probability 
algorithms) 
developed, among others, those in which diversification is 
time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
or diversity-dependent [\cite{etienne2011diversity}],
or that can shift [\cite{etienne2012conceptual}, 
\cite{rabosky2014automatic}, \cite{alfaro2009nine}, \cite{laudanno2018sls}].
\fi

Many of such models (and their associated probability 
algorithms) have been
developed to include diverse features: those in which diversification rates are 
time-dependent [\cite{nee1994reconstructed}, \cite{rabosky2008explosive}], 
or diversity-dependent [\cite{etienne2011diversity}],
or that can shift [\cite{etienne2012conceptual}, 
\cite{rabosky2014automatic}, \cite{alfaro2009nine}, \cite{laudanno2018sls}].

Other birth-death models treat speciation as a process that takes 
time [\cite{rosindell2010protracted}][\cite{etienne2012prolonging}], 
or that happens in bursts of simultaneous 
events [\cite{laudanno2018mbd}], or that
depends on a trait that has two [\cite{maddison2007estimating}], 
or more [\cite{fitzjohn2012diversitree}] states,
or even allows concealed states [\cite{beaulieu2016detecting}] 
or a combination of all these [\cite{herrera2018detecting}].
Only few of these diversification models are available as a BEAST2 package.

@@

Now we can formulate the first example research question that \verb;pirouette; 
can answer.

@@

@@

In this example we use a standard generative tree prior $\mathit{P_{0}}$, 
namely the Yule (pure-birth) tree prior. 
We choose to use a small tree with six taxa to keep
the calculations short and the figure more readable.
We pick a crown age of ten time units. This value is 
completely arbitrary, but it ties in with the mutation rate 
used in simulating an alignment in the next step.

@@

\begin{lstlisting}[
  language=R, floatplacement=ht, frame=single,
  label = {lst:create_error_measure_params},
  caption = Calling \texttt{create\_error\_measure\_params}.
]
error_measure_params <- create_error_measure_params(
  burn_in_fraction = 0.1, 
  error_function = get_nltt_error_function()
)
\end{lstlisting}

@@

\begin{lstlisting}[
  language=R, floatplacement=ht, frame=single,
  label = {lst:create_pir_params},
  caption = Calling \texttt{create\_pir\_params}.
]
pir_params <- create_pir_params(
  alignment_params = alignment_params,
  experiments = experiments,
  error_measure_params = error_measure_params,
  twinning_params = NA
)
\end{lstlisting}

@@

\begin{lstlisting}[
  language=R, floatplacement=ht, frame=single,
  label = {lst:pir_run},
  caption = Calling \texttt{pir\_run}.
]
errors <- pir_run(
  phylogeny = phylogeny,
  pir_params = pir_params
)
\end{lstlisting}

@@

\begin{lstlisting}[
  language=R, floatplacement=ht, frame=single,
  label = {lst:pir_plot},
  caption = Calling \texttt{pir\_plot}.
]
pir_plot(errors)
\end{lstlisting}

@@

The error distribution obtained is shown as a violin plot
in figure~\ref{fig:example_1}
\richel{
  I suggest to use a histogram (or density plot) instead:
  I like how the error distribution in figure 1 looks.
}
\giovanni{Can we show everything at the same time with standard histograms?}
.

@@

By doing so, this error distribution will serve as a control,
as it can be seen as the baseline error caused by stochasticity.
If the two error distributions obtained from $\mathit{P_{0}}$ and $\mathit{P}$ 
will appear similar, the error induced by the choice of a non-standard 
speciation process will be of little influence.
Given the necessity of such comparison,
we made it an inherent part of \verb;pirouette;. 
We call it 'twinning' and we use it from ERQ 3.

@@

If the user has sufficient reasons to believe that one standard tree prior 
could be more suitable than others for the inference, they can set it as part 
of the inference model along with site and clock models used to simulate the 
alignment. This option can be specified selecting "model\_type" to be 
"generative" as part of the inference conditions (as seen in ERQ 1, 
table~\ref{tab:RQ1}).

We show the procedure using \verb;pirouette; to answer the second research 
question.

@@

@@

This time we start with a tree generated from an unknown 
diversification model, that has six taxa and a crown age of ten:

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_2/errors.png}
  \caption{
    Example 2: the error BEAST2 makes from a phylogeny when
    picking the best inference model.
  }
  \label{fig:example_2}
  \giovanni{There is a weird NA at the top of the figure.}
\end{figure}

@@

This time, however, we are interested in measuring both the inference error
and a baseline error. We do so by setting the twinning parameters $\mathit{T}$ to create a twin tree. A twin tree
has the same topology as the given tree, yet its branching times are simulated 
according to the standard tree prior used by BEAST2 to produce the posterior 
for the main pipeline. 
Since in the twin pipeline there is coherence between the inference model and the generative model, 
we expect BEAST2 to produce a posterior of phylogenies that are more similar to the 
original one, respect to the outcome of the main pipeline. 
Due to this, the twinning process is used to provide a control for the main 
experiment.

@@

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_3/errors.png}
  \caption{
    Example 3: the error made by BEAST2 from a phylogeny, 
    when hand-picking a generative tree prior versus the control.
    Here, the 'twin' column shows the baseline inference error,
    as measured on the twin tree.
  }
  \label{fig:example_3}
\end{figure}

@@

To answer ERQ 2 we used an evidence-based model selection procedure to obtain 
an 
error distribution.
However, it is unclear to what extent this error is due to the mismatch between 
the tree priors (from generative and inference models) or due to stochasticity. 
As mentioned in ERQs 1 and 3, \verb;pirouette; provides a tool to evaluate this, using 
the twinning process. We now apply it to the same pipeline followed in ERQ 2.

@@

Here, as in ERQs 2 and 3, we start with a phylogeny 
generated by an unknown speciation model (listing~\ref{lst:unknown_phylogeny}). 
The first step in \verb;pirouette; is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment}.

@@

The output is shown in figure~\ref{fig:example_4}\giovanni{figure~\ref{fig:example_4} needs to be closer}. The alignment
generated for the true tree had most evidence for the JC site model,
strict clock model and Yule tree prior, 
with a weight of 0.981 (as 
shown in table \ref{tab:evidences_example_4}). 
The alignment
generated for the twin tree had most evidence for the JC site model,
RLN clock model and CBS tree prior,
with a weight of 0.707 (as 
shown in table \ref{tab:evidences_example_4}).

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_4/errors.png}
  \caption{
    Example 4: the error made by BEAST2 from a phylogeny, when the generative 
    model is known, compared to candidates with similar inference models, in 
    relationship to the background error.
  }
  \label{fig:example_4}
  \giovanni{NA on the top.}
\end{figure}

@@

In ERQs 1 and 2 we used \verb;pirouette; to  
run the analysis on the main pipeline, either using a 'generative' or a 
'best\_candidate' inference model, resulting in one error distribution,
that does not mean much on its own.
In ERQs 3 and 4 we measured the baseline error caused by stochasticity,
by running a parallel twin pipeline. This yielded two error distributions:
the baseline error and the complete inference error. Here we want to combine all the instances of the previous ERQs at once.

@@

In this example we use a tree generated by the Yule process.
Instead of using the common Yule tree as in example 1, we will use
a Yule tree of low (actually: zero) likelihood \giovanni{what do you mean by zero likelihood?}, i.e. the one already used in ERQs 2-4, by using the code shown in listing~\ref{lst:unknown_phylogeny}.
The first step in \verb;pirouette; is to simulate a DNA alignment 
from the given phylogeny, for which we will use the same code 
as shown in listing~\ref{lst:create_alignment}.

@@

In the second step we state our experiments. 
In this example we must specify both the generative and candidate experiments. 
Here we specify the generative model:

@@

We choose again to use the default error measurement parameters
created in listing~\ref{lst:create_error_measure_params}
and combine this with the same twinning parameters 
used in listing~\ref{lst:create_twinning_params}:

@@

The output is shown in figure~\ref{fig:example_5}.
\giovanni{figure~\ref{fig:example_5} needs to be closer}
The alignment
generated from the true tree had most evidence for the JC site model,
strict clock model and Yule tree prior, 
with a weight of 0.981 \richel{again} (as 
shown in table \ref{tab:evidences_example_5}). 
The alignment
generated for the twin tree had most evidence for the JC site model,
RLN clock model and CBS tree prior,
with a weight of 0.707 \richel{again} (as 
shown in table \ref{tab:evidences_example_5_twin}).

\begin{figure}[ht]
  \includegraphics[width=\textwidth]{example_5/errors.png}
  \caption{
    Example 5: the error made by BEAST2 from a phylogeny, 
    when the generative model is known, 
    compared to candidates with similar inference models, 
    in relationship to the background error.
  }
  \giovanni{na on the top-right area. labels on x superimpose each other. We should change the names to their shortened versions: e.g. relaxed\_log\_normal $\to$ rln}
  \label{fig:example_5}
\end{figure}
